### 2024-08-08 Article/main.tex
Preservation, reflection, lifting and creation of (co)limits

\begin{definition}[Limits]\label{def:limit}
    Let $D:\cat {I \rightarrow C}$ be a diagram in $\cat C$ of shape $\cat I$. A cone $\{f_i: X \rightarrow D(i)\}$ is a \emph{limit} provided that, for any other cone $\{f_{i}': X' \rightarrow D(i)\}$ for $D$, then there exists a unique morphism $k: X' \rightarrow X$ such that the following diagram commutes for each object $i$ of $\cat I$:
    \[
        \begin{tikzcd}
            X' \arrow[rr, dashed, "k"] \arrow[dr, "{f_i'}" swap] & & X \arrow[dl, "{f_i}"] \\
            & D(i) &
        \end{tikzcd}
    \]
    -- i.e., $f_i \circ k = f_i'$ for each object $i$ of $\cat I$.
\end{definition}

@@

    What we want now is to show that $(L, (p_i)_{i \in \Ob(\cat I)})$ is the limit of $D$, namely, show that conditions given at the beginning are valid.

@@

>>>>

\begin{definition}[Limits]\label{def:limit}
    Let $D:\cat {I \rightarrow C}$ be a diagram in $\cat C$ of shape $\cat I$. A cone $\{f_i: X \rightarrow D(i)\}$ is a \emph{limit} provided that, for any other cone $\{f_{i}': X' \rightarrow D(i)\}$ for $D$, then there exists a unique morphism $k: X' \rightarrow X$ such that the following diagram commutes for each object $i$ of $\cat I$:
    \[
        \begin{tikzcd}
            X' \arrow[rr, dashed, "k"] \arrow[dr, "{f_i'}" swap] & & X \arrow[dl, "{f_i}"] \\
            & D(i) &
        \end{tikzcd}
    \]
    -- i.e., $f_i \circ k = f_i'$ for each object $i$ of $\cat I$. Such limit is denoted as $(X, f_i)_{i \in \cat I}$
\end{definition}

@@

    What we want now is to show that $(L, (p_i))_{i \in \cat I}$ is the limit of $D$, namely, show that conditions given at the beginning are valid.

@@

In the next sections, we will work on a special kind of diagrams with certain properties. In particular, we are interested in how a functor behave whit respect to the constructions defined so far.

\begin{definition}
    Let $D : \cat{I \rightarrow C}$ be a diagram, and $F: \cat{C \rightarrow D}$ a functor. We say that $F$:
    \begin{enumerate}
        \item \emph{preserves limits} of $D$ if, given a limit $(L, l_i)_{i \in \cat I}$ for $D$, then $(F(L), F(l_i))_{i \in \cat I}$ is a limit for $F \circ D$.
        \item \emph{reflects limits} of $D$ if a cone $(L, l_i)_{i \in \cat I}$ is a limit for $D$ whenever $(F(L), F(l_i))_{i \in \cat I}$ is a limit for $F \circ D$.
        \item \emph{lifts limits} of $D$ if, given a limit $(L, l_i)_{i \in \cat I}$ for $F \circ D$, there exists a limit $(L', l_i')_{i \in \cat I}$ such that $(F(L'), F(l_i'))_{i \in \cat I} = (L, l_i)_{i \in \cat I}$.
        \item \emph{creates limits} of $D$ if $D$ has a limit and $F$ preserves and reflects limits along it.
    \end{enumerate}
    The dual notions are obtained in the obvious way, namely, substituting the words \textit{"limits"} and \textit{"cones"} with \textit{"colimits"} and \textit{"cocones"}.
\end{definition}
